---
output: pdf_document
---
# Use of Central Limit Theorem in Simulating Exponential Random Variables
## Data Analysis Series:  Statistical Inference
## Courtney D. Shelley
## August 21, 2014



###  Introduction

The exponential distribution describes the arrival time of a randomly recurring independent event sequence. If $\mu$ is the mean waiting time for the next event recurrence, its probability density function is:

   $$             
     P(x)=D'(x)=\lambda*e^{-\lambda x}  \mathbb{1}_{[0, \infty)}.
   $$ 
   

The exponential distribution can be simulated in R with rexp(n, lambda) where n is the number of lambda is the rate parameter. The mean of the exponential distribution is 1/lambda and the standard deviation is also also 1/lambda. 

###  Analysis

The purpose of this exercise is to illustrate via simulation the distribution of averages of 40 exponential (lambda = 0.2) simulations.  Approximately 1,000 simulated averages will be conducted.   

```{r createData}

####  1000 Simulations of the average of 40 exp(0.2) variables.  
means<- replicate(1000, mean(rexp(40, 0.2)))
```

### Goal 1:  Demonstrate the center of the distribution and compare to the theoretical center of the distribution.  

```{r meanAnalysis}
meanOfMeans <- mean(means)
expMean <- 1/0.2
```
> The simulated data mean is `r meanOfMeans`, which is very similar to the theoretical mean of 1/$\lambda$ or `r expMean`.  

###  Goal 2:  2. Demonstrate simulated distribution variation and compare to theoretical distribution.  
```{r varAnalysis}
varOfMeans <- var(means)
sdOfMeans <- sd(means)
expVar <- (1/0.2)^2
expSD <- 1/0.2
```
> The variation of the simulated data is `r varOfMeans` while the expected theoretical variation is `r expVar`.  It is not surprising that the simulated variation does not closely align with the theoretical variance, since the simulated means clearly follow a normal distribution themselves, as predicted by the Central Limit Theorem and demonstrated below.  

### Goal 3:  Demonstrate that the simulated means follow a normal distribution.  

Exploratory analysis of the simulated data by way of a histogram show that the simulated data are normally distributed about the mean = `r meanOfMeans`, with a standard deviation of `r sdOfMeans`.  
 
```{r histMeans}
Z<-(means-mean(means))/sdOfMeans
hist(Z, prob=TRUE, col="lightblue", breaks =  50, xlab = "Standardized Means", main="Simulated Means of 40 Simulated Exp(0.2)")                     
abline(v=0, col="red", lwd=3)
lines(density(Z), col="violet", lwd=2)              # add a density estimate with defaults
lines(density(Z, adjust=2), lty="dotted", col="black", lwd=2)   # add another smoother density
legend("topright", pch = 16, col = c("red", "violet", "black"), legend = c("Standardized Mean = 0", "Density Estimate", "Gaussian Approximation"))
```

### Goal 4:  Evaluate the coverage of the confidence interval for 1/lambda: $\bar{X}\pm1.96S/\sqrt{n}$.

```{r coverage}
ll<-mean(means)-1.96*expSD/sqrt(40)
ul<-mean(means)+1.96*expSD/sqrt(40)
coverage<-subset(means, means>ll & means<ul)
percent <- length(coverage)/1000*100
```
> The 95% confidence interval of simulated means is `r ll`-`r ul` with a `r percent`% coverage.  Since the method used to create the CI is low-biased, this coverage percentage makes sense.  